Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/overview.tex
working on the overview

The idea behind staged programming is to use staging annotations or
types so indicate the stage of each expression.  To apply this
technique to our example, suppose that we have a language with two
stages.  We can now write a staged version of \texttt{qs}, called
\texttt{qss} in \ref{fig:qs-staged}, making the input list a stage-1
and the rank a stage-2, more precisely the input list has type
$\curr\mathrm{list}$ (an integer list ``now''), the rank has type
$\fut\rmint$ (an integer in the ``future''), and return has type
$\fut\rmint$.  We can then annotate the rest of the body of
\texttt{qs}, by using the staging annotations \texttt{prev} and
\texttt{next}, which respectively include a stage-1 expressions in
stage-2, and a stage-2 expression in stage-1.  To help readability, we
write the first-stage computations in red, and the second-stage
computations in blue.

@@

We can formalize this intuition by writing quickselect in a \emph{staged}
language. In this paper, we choose \lang\ (\ref{sec:semantics}), a staged, typed
lambda calculus. This allows us to directly express the idea that the argument
\texttt{l} is known at the \emph{first} stage of the computation, while the
argument \texttt{k} is only known at the \emph{second}, and all
first-stage computations occur before any second-stage computations.

We define a staged version of quickselect in \ref{fig:qs-staged}, writing
first-stage computations in red, and second-stage computations in blue. While
\texttt{qSelect} sends a \textrm{list} and $\rmint$ to an $\rmint$,
\texttt{qsStaged} has a more precise type---it is a first-stage function which
takes a $\curr\mathrm{list}$ (an integer list now) and a $\fut\rmint$
(an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.

Ignoring for the moment all \texttt{grnd} annotations, the first-stage code in
\texttt{qsStaged} looks like \texttt{qSelect}: in the first stage, terms of
non-circle type are available for immediate use. Indeed, we case on the list as
usual, and in the \texttt{Cons} branch, \texttt{part}ition it.

@@

The process of adding staging annotations ($\fut$ types, $\next$, and $\prev$)
to unstaged code has been the subject of extensive research under the
name of \emph{binding time analysis}. In this paper, we assume that these
annotations have been provided by the programmer (or perhaps a binding time
analysis tool), and do not consider the problem of generating such annotations.
There are many ways to annotate any program, including \texttt{qSelect}; we
chose annotations which maximize the work performed in the first stage.

\lang's type system, discussed in \ref{sec:semantics}, ensures that the staging
annotations are consistent, in the sense that computations marked as
first-stage cannot depend on ones marked as second-stage. The $\curr$ and
\texttt{grnd} (``ground'') annotations further distinguish those first-stage computations
which do not contain any embedded second-stage code. (Code \emph{inside}
\texttt{grnd}, like the \texttt{part}ition function, is guaranteed not to
contain second-stage code; first stage code outside a \texttt{grnd}, 
like \texttt{qsStaged}, may.) We will
discuss the importance of this additional distinction in \ref{sec:splitting}.

>>>>

The idea behind staged programming is to use staging annotations or
types to indicate the stage of each expression.  To apply this
technique to our example, suppose that we have a language with two
worlds representing the two stages and a third, {\em ground} world
that contains terms with no staging.

We can write a staged version of \texttt{qs}, called
\texttt{qss} in \ref{fig:qs-staged}, by making the input list a
stage-1 and the rank a stage-2 argument. 
%
More precisely the input list
has type $\curr\mathrm{list}$ (an integer list ``now''), the rank has
type $\fut\rmint$ (an integer in the ``future''), and return has type
$\fut\rmint$.  
%
We can then obtain \texttt{qss} by annotating the body of \texttt{qs}
with the staging annotations \texttt{prev} and \texttt{next}, which
transition between a stage-1 and stage-2 code, and $\texttt{g}$, which
marks ground terms.  
%
We also use a third function
\texttt{hold}, which can be implemented with \texttt{prev} and
\texttt{next}, for promoting a stage-1 integer to a stage-2 integer;
the signature of \texttt{hold} is
% 
\lstinline{1`hold : ^int -> $`2`int`}.

To help readability, we write the first-stage
computations in red, and the second-stage computations in blue.

Our type system ensures that the staging annotations are consistent,
in the sense that computations marked as first-stage cannot depend on
ones marked as second-stage.
%
The process of adding staging annotations to unstaged code has been
the subject of extensive research (\secref{related}). We assume that
these annotations have been provided; we do not consider the problem
of generating them. In this example, there are other ways to annotate
\texttt{qs}, but we chose annotations that maximize the work performed
in the first stage

% The $\curr$ and \texttt{grnd} (``ground'') annotations
% further distinguish those first-stage computations which do not
% contain any embedded second-stage code. (Code \emph{inside}
% \texttt{grnd}, like the \texttt{part}ition function, is guaranteed not
% to contain second-stage code; first stage code outside a
% \texttt{grnd}, like \texttt{qsStaged}, may.) We will discuss the
% importance of this additional distinction in \ref{sec:splitting}.

% Ignoring for the moment all \texttt{grnd} annotations, the first-stage code in
% \texttt{qsStaged} looks like \texttt{qSelect}: in the first stage, terms of
% non-circle type are available for immediate use. Indeed, we case on the list as
% usual, and in the \texttt{Cons} branch, \texttt{part}ition it.

@@

@@
